
## Regression and model validation

```{r}
date()
```
**FIRST, I carried out the data wrangling exercise and created the following R script.**

*#Ari Voutilainen #November 9, 2021. #R script for data wrangling in IODS2021 Exercise 2.*

- lrn14 <- read.table("http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-data.txt", sep="\t", header=TRUE)
- dim(lrn14)
- str(lrn14)

*#LRN14 consists of 183 rows and 60 columns. #The last four columns are entitled as follows: Age, Attitude, Points, and gender.*

*#Creating variables deep, stra, surf, and atti.*

- library(dplyr)
- deep_questions <- c("D03", "D11", "D19", "D27", "D07", "D14", "D22", "D30","D06",  "D15", "D23", "D31")
- deep_columns <- select(lrn14, one_of(deep_questions))
- lrn14$deep <- rowMeans(deep_columns)
- strategic_questions <- c("ST01","ST09","ST17","ST25","ST04","ST12","ST20","ST28")
- strategic_columns <- select(lrn14, one_of(strategic_questions))
- lrn14$stra <- rowMeans(strategic_columns)
- surface_questions <- c("SU02","SU10","SU18","SU26", "SU05","SU13","SU21","SU29","SU08","SU16","SU24","SU32")
- surface_columns <- select(lrn14, one_of(surface_questions))
- lrn14$surf <- rowMeans(surface_columns)
- lrn14$atti <- lrn14$Attitude / 10

*#I found the name of the one_of function misleading. It does not select just one but all.*

*#Creating an analysis data set.*

- keep_columns <- c("gender","Age","atti", "deep", "stra", "surf", "Points")
- learning2014 <- select(lrn14, one_of(keep_columns))
- learning2014 <- filter(learning2014, Points > 0)

*#Now the analysis data set consists of 166 rows and 7 columns.*

*#Setting the working directory.*

- setwd("C:/Users/ariv/Documents/IODS-project/IODS-project/data")
- getwd()

*#Obs! There is an IODS-project folder inside an IODS-project folder on purpose.*

*#Saving the analysis data set.*

- ?write.csv
- write.csv(learning2014, file = "learning2014.csv")
- read.csv("learning2014.csv")
- str(learning2014)
- head(learning2014)

**SECOND, I explored the structure of the learning2014 data set.**

- setwd("C:/Users/ariv/Documents/IODS-project/IODS-project/data")
- read.csv("learning2014.csv")
- dim(learning2014)
- str(learning2014)
- head(learning2014)

The **learning2014** data set describes students in the Introduction to Social Statistics course in 2014. The number of students is 166, and all of them participated in the course exam.

The **learning2014** data set consists of seven variables:

1. *Gender*.
2. *Age* in years.
3. Global *attitude* toward statistics.
4. Mean answer to questions reflecting *deep learning*.
5. Mean answer to questions reflecting *strategic learning*.
6. Mean answer to questions reflecting *surface learning*.
7. *Points* achieved in the exam.

**THIRD, I explored the variables of the learning2014 data in more detail.**

```{r}
library(readr)
setwd("C:/Users/ariv/Documents/IODS-project/IODS-project/data")
learning2014 <- read_csv("learning2014.csv", 
    col_types = cols(...1 = col_skip(), gender = col_factor(levels = c("F", 
        "M")), Age = col_number(), atti = col_number(), 
        deep = col_number(), stra = col_number(), 
        surf = col_number(), Points = col_number()))
View(learning2014)
summary(learning2014$gender)
summary(learning2014$Age)
summary(learning2014$atti)
summary(learning2014$deep)
summary(learning2014$stra)
summary(learning2014$surf)
summary(learning2014$Points)
```

Of the 166 students, 110 were female. The mean (range) age of students was 25.5 (17-55) years. Their global attitude toward statistics ranged from 1.4 to 5.0 with a mean value of 3.1. Variables describing deep, strategic, and surface learning resulted in the following mean (range) values, respectively: 3.7 (1.6-4.9), 3.1 (1.3-5.0), and 2.8 (1.6-4.3). In the variables reflecting attitude and learning strategies, the scale ranged from 1 to 5.

```{r}
pairs(learning2014[-1])
```

Based on a visual inspection of the scatter plot matrix, exam points correlated most evidently with students' attitude.
