
## Analysis of longitudinal data

```{r}
date()
```
Before the analyses, I created a new R Markdown file, named it as 'chapter6.Rmd', and saved it to my folder. Then, I included the new file as a child file in my 'index.Rmd' file.

**FIRST, I implemented the analyses of Chapter 8 of MABS using the RATS data.**

```{r echo=FALSE}
library(readr)
RATSL <- read_csv("RATSL.csv")
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
summary(RATSL)
```
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

Individual response profiles of **non-standardized weights** by Group highlight differences across Groups. Weights of rats belonging to Group 1 are lower in general, also at baseline.

```{r}
RATSL <- RATSL %>%
  group_by(Group) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "Standardized Weight")
```

Individual response profiles of **standardized weights** emphasize differences across individuals.

```{r}
n <- 16
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(stdweight), se = sd(stdweight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.1,0.8)) +
  scale_y_continuous(name = "Mean(StdWeight) +/- SE(StdWeight)")
```

Mean response profiles for the three treatment groups show no differences across diets with respect to standardized weight gain.

