
## Logistic regression and cross-validation

```{r}
date()
```
**FIRST, I created a new R Markdown file**, named it as 'chapter3.Rmd', and saved it to my folder. Then, I included the new file as a child file in my 'index.Rmd' file.

**SECOND, I read the joined and modified 'alc' data into R from my local folder, and described it briefly.**

```{r}
library(readr)
setwd("C:/Users/ariv/Documents/IODS-project/IODS-project/data")
alc <- read_csv("alc.csv")
View(alc)
```

```{r}
colnames(alc)[2:36]
```

The **alc** data set describes students of two Portuguese schools, their age, gender, socio-demographics, free-time including alcohol use, health status, etc. The data set consists of 370 students and 35 variables. Prof. **Paulo Cortez** (http://www3.dsi.uminho.pt/pcortez/Home.html) is the provider of the original data set (**Student Performance Data Set**, https://archive.ics.uci.edu/ml/datasets/Student+Performance).

**THIRD, I selected four potential explanatory variables for alcohol consumption** from the **alc** data set, explored them, and hypothesized their relationships with alcohol consumption.

*This corresponds to the data analysis part of Exercise 3, its 3. and 4. tasks.*

The first potential explanatory variable for alcohol consumption was **sex**, as, in general, there are gender differences in alcohol consumption (Wilsnack et al. 2000 https://onlinelibrary.wiley.com/doi/abs/10.1046/j.1360-0443.2000.95225112.x). I hypothesized that males consume more alcohol and are more often heavy drinkers than females.

```{r}
library(dplyr); library(ggplot2); library(readr)
setwd("C:/Users/ariv/Documents/IODS-project/IODS-project/data")
alc <- read_csv("alc.csv")
alc %>% group_by(sex, high_use) %>% count
```

The number of males in the **alc** data is 175, and that of females is 195. Of the 175 males, 70 i.e. 41% are heavy drinkers, whereas of the 195 females only 41 i.e. 21% are heavy drinkers, which appears to support the hypothesis.

The second potential explanatory variable for alcohol consumption was **age**. I expected that the proportion of heavy drinkers increases together with age.

```{r}
alc %>% group_by(high_use) %>% summarise(count = n(), mean_age = mean(age))
g1 <- ggplot(alc, aes(x = high_use, y = age))
g1 + geom_boxplot() + xlab("Heavy drinking") + ylab("Age in years") + ggtitle("Heavy and non-heavy drinking students by age")
```

The mean age of heavy drinking students was 16.8 years and that of non-heavy drinking students was 16.5 years. Based on the box plot the median age of heavy-drinking students was 17 years and that of non-heavy drinking students was 16 years. Thus, the age distribution slightly suggested that heavy-drinking students are older.

The third potential explanatory variable for alcohol consumption was **goout** that indicates going out with friends as a numeric values from 1 (very low frequency) to 5 (very high frequency). By and large, friends' drinking behavior is related to adolescents' drinking (Bot et al. 2005 https://www.sciencedirect.com/science/article/pii/S0306460304003314) and, consequently, I expected going out with friends associates with heavy-drinking also in the *alc* data.

```{r}
g2 <- ggplot(alc, aes(x = high_use, y = goout))
g2 + geom_boxplot() + xlab("Heavy drinking") + ylab("Going out intensity") + ggtitle("Heavy and non-heavy drinking students by going out with friends")
```

The distribution of the **goout** variable suggested that heavy drinking students go out with friends more frequently and/or with higher intensity than non-heavy drinking students. This supports my hypothesis.

The fourth potential explanatory variable for alcohol consumption was **address** that indicates living in an urban (U) or rural area (R). I chose this variable because in some European countries, such as Germany, adolescents living in rural areas appear to be more susceptible to drinking problems (Donath et al. 2011 https://bmcpublichealth.biomedcentral.com/articles/10.1186/1471-2458-11-84).

```{r}
alc %>% group_by(address, high_use) %>% count
```

In the **alc** data, the number of students living in rural areas was 81 and that of students living in urban areas was 289. Of the 81 'rural students', 31 i.e. 38% was heavy drinkers, whereas of the 289 'urban students' only 28% were heavy drinkers. This supports my hypothesis that living in a rural area predispose students to drinking problems.

**FOURTH, I performed logistic regression** to explore the relationship between the four selected explanatory variables (**sex, age, goout, address**) and alcohol consumption. As a target variables, I used the binary **high_use** variable.

```{r}
m <- glm(high_use ~ sex + age + goout + address, data = alc, family = "binomial")
summary(m)
coef(m)
OR <- coef(m) %>% exp
CI <- confint(m) %>% exp
cbind(OR, CI)
```

Based on the logistic regression model with **sex**, **age**, **goout**, and **address** as explanatory variables, male gender, going out with friends, and living in a rural area increased the odds of being a heavy drinker. The odds of **men** was 2.5-times the odds of women (OR, 95% CI: 2.51, 1.54-4.15). One point increase in the **going out** scale increased the odds by 2.2 (OR, 95% CI: 2.16, 1.71-2.77). The odds of students living in an **urban** area was 0.5-times the odds of students living in a rural area (OR, 95% CI: 0.52, 0.29-0.94). With respect to age, the 95% CI for the OR contained 1, which denotes that age had no statistically significant effect on the OR. The result supported my hypotheses except for age.

**FIFTH, I explored the predictive power of the model.**

```{r}
m <- glm(high_use ~ sex + age + goout + address, data = alc, family = "binomial")
probabilities <- predict(m, type = "response")
alc <- mutate(alc, probability = probabilities)
alc <- mutate(alc, prediction = probability >0.5)
select(alc, sex, age, goout, address, high_use, probability, prediction) %>% tail(10)
table(high_use = alc$high_use, prediction = alc$prediction)
```

The **training error** of the model i.e. the total proportion on inaccurately classified individuals was (16+65)/(243+16+65+46)*100 = 21.9%. The model was better in predicting non-heavy drinking students. Of the 111 heavy-drinkers the model 'correctly recognized' only 46 (41%), whereas of the 259 non-heavy drinkers the model 'correctly recognized' 243 (94%).

```{r}
library(dplyr); library(ggplot2)
g <- ggplot(alc, aes(x = probability, y = high_use, col = prediction))
g + geom_point()
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table %>% addmargins
```

To compare the model performance to guessing I created a **guesspred** variable based on random numbers.

```{r}
alc$guessprop <- runif(370)
alc$guesspred <- ifelse(alc$guessprop > 0.5, "TRUE","FALSE")
table(high_use = alc$high_use, prediction = alc$guesspred)
table(high_use = alc$high_use, prediction = alc$guesspred) %>% prop.table %>% addmargins
```

In the guessing example, the **training error** was (124+54)/(135+124+54+57)*100 = 48.1% being much higher compared to the actual model.

**SIXTH, I carried out the BONUS task and performed 10-fold cross-validation on my model.**

```{r}
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}
loss_func(class = alc$high_use, prob = alc$probability)
library(boot)
cv <- cv.glm(data = alc, cost = loss_func, glmfit = m, K = 10)
cv$delta[1]
```

In my model, the original prediction error was 21.9%. In the 10-fold cross-validation, the mean prediction error increased to 25.1% still being a slightly lower than that of the DataCamp model (25.9%). Consequently, **sex**, **age**, **goout**, and **address** appear to form a bit better combination of explanatory variables than those used in the DataCamp model.

OBS! I did not carried out the SUPER-BONUS task.
