
## Analysis of longitudinal data

```{r}
date()
```
Before the analyses, I created a new R Markdown file, named it as 'chapter6.Rmd', and saved it to my folder. Then, I included the new file as a child file in my 'index.Rmd' file.

**FIRST, I implemented the analyses of Chapter 8 of MABS using the RATS data.**

```{r echo=FALSE}
library(readr)
RATSL <- read_csv("RATSL.csv")
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
summary(RATSL)
```
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

Individual response profiles of **non-standardized weights** by Group highlight differences across Groups. Weights of rats belonging to Group 1 are lower in general, also at baseline.

```{r}
RATSL <- RATSL %>%
  group_by(Group) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "Standardized Weight")
```

Individual response profiles of **standardized weights** emphasize differences across individuals.

```{r}
n <- 6 #In Group 1 n = 8, Group 2 n = 4, and Group 3 n = 3.
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(stdweight), se = sd(stdweight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.1,0.8)) +
  scale_y_continuous(name = "Mean(StdWeight) +/- SE(StdWeight)")
```

Mean response profiles for the three treatment groups reveal no difference in weight gain across diets.

```{r}
RATSL8S <- RATSL %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(stdweight) ) %>%
  ungroup()
glimpse(RATSL8S)
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "Mean(StdWeight), WD1-WD64")
```

Box plots of standardized mean summary weights by the three treatment groups suggest one outlier per Group.

```{r}
RATSL8S1 <- RATSL8S %>%
  filter(between(mean, -1, 1))
glimpse(RATSL8S1)
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "Mean(StdWeight), WD1-WD64")
```

After removing the outliers, the box plots indicate difference in weight gain across diets. Group 2 has gained less weight.

```{r}
summary(aov(mean ~ Group, data = RATSL8S1, var.equal = TRUE))
```

One-Way ANOVA verifies the box plot; there is a statistically significant difference in weight gain across diets after removing three outliers. The test P-value for Group is ~0.010.

```{r}
RATS <- read_csv("RATS.csv")
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)
anova(fit)
```

Using the first weight values (WD1) as covariates in the ANOVA does not change the interpretation. There still is a statistically significant difference in weight gain across diets. The test P-value for Group is ~0.002.

**SECOND, I implemented the analyses of Chapter 9 of MABS using the BPRS data.**

Before plotting the data and running the analyses, I added unique IDs for the subjects. In the original data, the **subject** variable does not distinguish the subjects by the treatment, only within the treatments.

```{r}
library(readr)
BPRSL <- read_csv("data/BPRSL.csv", col_types = cols(treatment = col_factor(levels = c("1", 
    "2")), subject = col_factor(levels = c("1", 
    "2", "3", "4", "5", "6", "7", "8", "9", 
    "10", "11", "12", "13", "14", "15", "16", 
    "17", "18", "19", "20"))))
ID2 <- c(21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40)
BPRSL <- mutate(BPRSL, ID = ifelse(treatment==1, subject,ID2))
BPRSL$ID <- factor(BPRSL$ID)
names(BPRSL)
summary(BPRSL)
glimpse(BPRSL)
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
dim(BPRSL)
ggplot(BPRSL, aes(x = week, y = bprs, group = treatment)) + geom_point(aes(colour = treatment), size = 2)
BPRSS <- BPRSL %>%
  group_by(treatment, week) %>%
  summarise( mean = mean(bprs) ) %>%
  ungroup()
ggplot(BPRSS, aes(x = week, y = mean, linetype = treatment, shape = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "BPRS")
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
ggplot(BPRSL, aes(x = week, y = bprs, group = ID)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Week", breaks = seq(0, 8)) +
  scale_y_continuous(name = "BPRS") +
  theme(legend.position = "top")
library(readr)
BPRS <- read_csv("BPRS.csv")
pairs(BPRS[4:12])
```

Based on plots, there is no difference in BPRS between the treatments, but there is substantial between-subject variation, and time remarkably affects BPRS. Moreover, time points appear to correlate with each other.

```{r}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```

The multiple linear regression model supports the interpretation based on plots. Week has a statistically significant effect on BPRS, the P-value is extremely low, whereas Treatment has no effect (P-value >>0.05).

```{r}
library(lme4)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | ID), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | ID), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref1, BPRS_ref)
```

```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | ID), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
anova(BPRS_ref2, BPRS_ref1)
```
Altogether, I performed four models regarding effects of two treatments on BPRS among men. The models were: the independence model (a multiple liner regression), the random intercept (RI) model, the random intercept and random slope (RIS) model, and the random intercept and random slope model with interaction (RISI). 

All models resulted in the same conclusion: There were no statistically significant differences between the treatments, but BPRS decreased significantly over the study.

Based on AIC, BIC, and the likelihood ratio test the RIS model provided a better fit for the BPRS data than the RI model, but there were no difference in the model fit between the RIS and RISI models. Actually, based on AIC and BIC the RIS model was even narrowly better.

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, group = ID)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Week", breaks = seq(0, 8)) +
  scale_y_continuous(name = "BPRS") +
  theme(legend.position = "top")
Fitted <- fitted(BPRS_ref2)
BPRSL <- BPRSL %>% mutate(Fitted)
ggplot(BPRSL, aes(x = week, y = Fitted, group = ID)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Week", breaks = seq(0, 8)) +
  scale_y_continuous(name = "Fitted BPRS") +
  theme(legend.position = "top")
```

To produce the fitted BPRS I used the RISI model. compared to the observed BPRS the fitted profiles are more straightforward and them simplify the entity, but, all in all, the fitted model was able to capture the process.