
## Clustering and classification

```{r}
date()
```
**FIRST, I created a new R Markdown file**, named it as 'chapter4.Rmd', and saved it to my folder. Then, I included the new file as a child file in my 'index.Rmd' file.

**SECOND, I load the Boston data set** from the MASS package and explored its structure and dimensions.

```{r}
library(MASS)
data("Boston")
str(Boston)
summary(Boston)
```

The **Boston** data consist of 506 rows and 14 columns. It describes demographic, socioeconomic, and environmental characteristics of towns.

References to the data source are as follows:

- Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81–102.
- Belsley D.A., Kuh, E. and Welsch, R.E. (1980) Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley.

**THIRD, I graphically and numerically described the Boston data** variables and relationships across them.

```{r}
library(corrplot)
library(tidyr)
pairs(Boston)
cor_matrix<-cor(Boston) %>% round(digits = 2)
cor_matrix
```

