
## Clustering and classification

```{r}
date()
```
**FIRST, I created a new R Markdown file**, named it as 'chapter4.Rmd', and saved it to my folder. Then, I included the new file as a child file in my 'index.Rmd' file.

**SECOND, I loaded the Boston data set** from the MASS package and explored its structure and dimensions.

```{r}
library(MASS)
data("Boston")
str(Boston)
summary(Boston)
```

The **Boston** data consist of 506 rows and 14 columns. It describes demographic, socioeconomic, and environmental characteristics of towns.

References to the data source are as follows:

- Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81–102.
- Belsley D.A., Kuh, E. and Welsch, R.E. (1980) Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley.

**THIRD, I graphically and numerically described the Boston data** variables and relationships across them.

```{r echo=FALSE}
library(corrplot)
library(tidyr)
library(MASS)
data("Boston")
par(mfrow = c(3,5))
hist(Boston$crim, main="crim", xlab="")
hist(Boston$zn, main="zn", xlab="")
hist(Boston$indus, main="indus", xlab="")
hist(Boston$chas, main="chas", xlab="")
hist(Boston$nox, main="nox", xlab="")
hist(Boston$rm, main="rm", xlab="")
hist(Boston$age, main="age", xlab="")
hist(Boston$dis, main="dis", xlab="")
hist(Boston$rad, main="rad", xlab="")
hist(Boston$tax, main="tax", xlab="")
hist(Boston$ptratio, main="ptratio", xlab="")
hist(Boston$black, main="black", xlab="")
hist(Boston$lstat, main="lstat", xlab="")
hist(Boston$medv, main="medv", xlab="")
```

Most **Boston** data variables are not normally distributed. By eye, only **rm** i.e. the average number of rooms per dwelling appears to follow a normal distribution.

```{r echo=FALSE}
library(corrplot)
library(tidyr)
library(MASS)
data("Boston")
pairs(Boston)
cor_matrix<-cor(Boston) %>% round(digits = 2)
cor_matrix
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6)
```

There are some strong correlations (*r* >0.70) across the **Boston** data variables. The strong positive correlations are **tax** x **rad** (*r* = 0.91), **nox** x **indus** (*r* = 0.76), **nox** x **age** (*r* = 0.73), and **tax** x **indus** (*r* = 0.72). Correspondingly, the strong negative correlations are **nox** x **dis** (*r* = -0.77), **age** x **dis** (*r* = -0.75), **medv** x **lstat** (*r* = -0.74), and **dis** x **indus** (*r* = -0.71).
