
## Dimensionality reduction techniques

```{r}
date()
```
**FIRST, I created a new R Markdown file**, named it as 'chapter5.Rmd', and saved it to my folder. Then, I included the new file as a child file in my 'index.Rmd' file.

**SECOND, I numerically and graphically described the human data** variables and relationships across them.

```{r echo=FALSE}
library(readr)
setwd("C:/Users/ariv/Documents/IODS-project/IODS-project/data")
human <- read_csv("human.csv")
summary(human)
```

```{r echo=FALSE}
library(corrplot)
library(tidyr)
par(mfrow = c(2,4))
hist(human$Edu2.FM, main="Edu2.FM", xlab="")
hist(human$Labo.FM, main="Labo.FM", xlab="")
hist(human$Edu.Exp, main="Edu.Exp", xlab="")
hist(human$Life.Exp, main="Life.Exp", xlab="")
hist(human$GNI, main="GNI", xlab="")
hist(human$Mat.Mor, main="Mat.Mor", xlab="")
hist(human$Ado.Birth, main="Ado.Birth", xlab="")
hist(human$Parli.F, main="Parli.F", xlab="")
```

Most **human** data variables are not normally distributed. By eye, only **Edu.Exp** i.e. the expected years of schooling appears to follow a normal distribution.

```{r echo=FALSE}
cor(human[,c('Edu2.FM','Labo.FM','Edu.Exp','Life.Exp','GNI','Mat.Mor','Ado.Birth','Parli.F')])
```

There are a few strong correlations (*r* larger than 0.70 or smaller than -0.70) across the **human** data variables. The strong positive correlations are **Edu.Exp** x **Life.Exp** (*r* = 0.79) and **Mat.Mor** x **Ado.Birth** (*r* = 0.76). Correspondingly, the strong negative correlations are **Life.Exp** x **Mat.Mor** (*r* = -0.86), **Edu.Exp** x **Mat.Mor** (*r* = -0.74), and **Life.Exp** x **Ado.Birth** (*r* = -0.73).

**THIRD, I performed the principal component analysis (PCA) on the non-standardized human data.**

```{r}
pca_human1 <- prcomp(human[-1])
biplot(pca_human1, choices = 1:2, cex = c(0.6, 0.6), col = c("grey40", "deeppink2"))
```

**FOURTH, I performed the principal component analysis (PCA) on the standardized human data.**

```{r}
human_std <- scale(human[-1])
pca_human2 <- prcomp(human_std)
biplot(pca_human2, choices = 1:2, cex = c(0.6, 0.6), col = c("grey40", "deeppink2"))
```

Text...